Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-06-18 manuscript/manuscript.tex
Cleanup after mistake with github.
First attempt to reducing size of the document
Also, updated post_closed_violin chart (color change) which is reflected in datacleansing.R also.

\begin{itemize}
	\item NYC restaurant violations
	\item Popular baby names
	\item Mapping of car crashes involving pedestrians
	\item Mapping of sidewalk widths in NYC 
	\item Visualization of NYC High School and College enrollment
	\item An interactive dashboard to filter and view charges for jail inmates
	\item Location of City-wide free Internet access points
	\item ...and 220 other complaint types. 
\end{itemize}

@@

The paper is organized as follows:
Section~\ref{sec:history} offers a brief review of the history of the 311 system. In 
Section~\ref{sec:trends} we take a look at long-term trends presented via a 10-year analysis.
Section~\ref{sec:issues} offers a general discussion of data cleansing issues
impacting data quality and curation efficiency. Section~\ref{sec:structural} examines
the technical dimensions of the dataset, the structural issues. 
In Section~\ref{sec:datatypes} we examine the data fields for compliance 
with the stated data type as found in the Data Dictionary. Section~\ref{sec:blanks} looks
at the data fields as regards missing, blank, or N/A entries. Section~\ref{sec:domain} explores
how data fields comply with a domain of legal or acceptable values. Section~\ref{sec:inconsistencies}
deals with the important issues surrounding logical inconsistencies 
and concerning patterns in the data. Section~\ref{sec:precision} 
explores the age-old issue of precision versus accuracy. Section~\ref{sec:duplicates}
identifies duplicate and redundant data fields. Section~\ref{sec:improvements} provides 
actionable suggestions for mitigating or resolving identified issues. 
Following this, Section~\ref{sec:protocol} outlines a series of general 
principles for the release of open data, drawing from our findings. The 
paper concludes with a discussion and recommendations in 
Section~\ref{sec:discussion}, encapsulating the key insights and 
implications of our research.

A note on naming conventions: dataset field names are typically 
named using ``snake\_case'' where each word in the variable name 
is written in lowercase, and words are separated by 
underscores, e.g. created\_date, complaint\_type, zip\_code, etc..

\section{NYC 311 System History} \label{sec:history}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. The evolution of this system can be traced from
its initial implementation as a simple inquiry channel via telephone to a
comprehensive data management system that handles millions of requests
annually. Each inquiry is logged as a complaint in the 311 system.
Key milestones in the 311 system development include:

\begin{itemize}
 	\item 2003 - NYC 311 system goes live with a phone-based call center only. 
 	The ``311'' phone number replaces a myriad of numbers to 
 	call for each individual City Agency.
   	\item 2009 - 311 online \& mobile apps launched. Today 60\% of all 
   	Service Requests come from mobile \& online channels.
   	\item 2019 - Major software upgrade of 311 system with enhanced capabilities, 
   	dynamic load-balancing, and cloud-based resilience. Only major 
   	software upgrade since 2004.	
      \item 2020 - In August 2020, driven by the COVID pandemic, monthly 311 
      Service Requests hit an all-time high of 348,463. Seven of the Top 10 busiest 
      days of all-time occur in 2020.
      \item 2021 - 311 System is expanded to into the MTA's city subway system, 
      the largest expansion since its inception.
   	\item 2023 - Record high year for 311 with 3.23 million Service Requests.
\end{itemize}

Today, the NYC 311 data system is a robust platform that manages a
vast array of urban living-related inquiries. The system handles over 3 million service 
requests (SRs) per year, encompassing such complaints as street noise, 
illegal parking, heat/hot water, abandoned vehicles,
and unsanitary conditions. The data, managed through a tailored application built on the 
Microsoft Dynamics and Azure platform, is a valuable resource for city
administration and policy-making. The 311 data is publicly accessible through the 
NYC Open Data Portal which provides access
to all the data sets (approx. 2700) and provides a convenient method for querying, 
grouping, aggregating, geo-mapping, visualization, and exporting 
results. This open data initiative enables not only governmental transparency 
but also empowers researchers, civic developers, and the general public.

Additional Open Data projects can be found 
at \href{https://opendata.cityofnewyork.us/projects/}{NYC Open Data Project Gallery}
and at \href{https://beta.nyc/beta/products/}{\textbeta etaNYC} products and tools. 
Despite its success, the system constantly faces challenges such as:

\begin{itemize}
	\item Data timeliness, accuracy, and consistency
	\item Difficulties correlating data over long time periods, e.g. over a 10-year 
	period, e.g. Agency name changes, new complaint types.
	\item Data anonymization and handling of Personally identifiable information (PII)
	\item Integration with stand-alone systems at selected NYC Agencies
	\item Managing API usage, authentication, and load for 3\textsuperscript{rd} 
	party users, of which there are many
	\item Incorporating ongoing technology changes and upgrades 
	\item Understanding the role of chatbots and AI to provide online 
	customer service, along with other emergent technologies
\end{itemize}

These and other challenges are continually addressed by the various 
Agency open data managers as well as the 
\href{https://www.nyc.gov/content/oti/pages/}
{NYC Office of Technology and Innovation (OTI)} which provides the 
technology support for the open data system. Many of the underlying 
datasets are updated daily, usually running 1-2 days behind 
from creation. However, a number of datasets are updated 
monthly (such as restaurant inspections), and some are not 
updated for one or more years (such as community board 
boundaries). This poses problems when merging datasets with 
others, as would be expected. Typically, the data must be exported 
and manipulated in another, external program. The authors found 
this to be the case with the large 311 dataset employed.

Data accuracy and consistency is typically established between OTI 
and the other contributing City Agencies, such as NYPD, Housing 
Preservation and Development (HPD), Parks \& Recreation, etc., via 
Service Level Agreements (SLAs). However, enforcement of those 
SLAs is often challenging. The result is that often data accuracy and 
logical consistency is occasionally compromised. However, 
the Open Data Team is quite responsive to inquires from users. The 
authors have provided many such suggestions during the course of this research effort, 
and the most current data shows improvements in many of the areas highlighted. 
Questions and suggestion  can be provide here: 
\href{https://opendata.cityofnewyork.us/engage/}{Contact Us}.

The NYC 311 organization, part of OTI, operates a highly tailored and sophisticated case
management. However, one constraint was particularly challenging 
during the recent complete overhaul of the software and hardware in 2019. When the original 311system was developed in 2003, a decision was made to allow select 
NYC Agencies to maintain their own local computer systems, 
such as the mainframe systems at the NYC Department of Sanitation (DSNY) and others. 
And when the 311 system was completely upgraded in 2019, those
same integrations were honored in order to reduce the scope of the project. 
Perhaps in retrospect, while a good budgetary and policy decision, 
it was perhaps not such a good technology decision.

At least nine NYC Agencies have local systems that are integrated 
directly to the 311 system either via an API or custom code. These include:

\begin{itemize}
	\item Department of Social Services (DSS), one of the largest City Agencies
	\item Department of Sanitation, although recently part of DSNY 
	has moved to the core 311 software.
	\item Department of Parks \& Recreation (forestry complaints only)
	\item Housing, Preservation \& Development (fully integrated)
	\item Department of Environmental Protection (DEP) (fully integrated)
	\item Department of Healthy \& Mental Hygiene (fully integrated) 
	\item Department of Consumer \& Worker Protection 
	(DCWP - formerly known as the Dept of Consumer Affairs)
	\item New York Police Department (NYPD) (integrated with their reporting system)
	\item Department of Business (DoB) (fully integrated)
\end{itemize}

One unfortunate side-effect from these integrations, and other Agencies 
that use the 311 APIs, is that some of the data consistency
checks which are inherent in the 311 system are not applied uniformly, 
and in some cases not at all, to the integrated data streams.
Errors often creep in from other Agency systems and are not easily 
corrected or detected in the 311 application itself,
but rather require correction at the source Agency; challenging.

One area that will almost certainly remain an ongoing issue is the 
ongoing incorporation of new software, hardware,
and cloud services which are constantly changing while new approaches emerge. 
The 311 telephonic integration recently was reconfigured due to a new
telephonic vendor. Desktop software at the 311 Agency are subject 
to the same upgrades that home and business users experience. And while the 
311 system was designed to be easily configured and updated with new features, 
some emerging business issues as well as technologies nonetheless 
require additional software engineering efforts to accommodate, 
along with the concomitant budgets.

One area in particular which is heavily influencing customer outreach for the 311 system
 is the appropriate use of online chatbots and assistants. Increasing these features are 
 influenced by artificial intelligence (AI) advances. It is a given that a number of 
 citizen's interactions with City government could be handled accurately and quickly by AI-driven
applications thus improving timeliness and accuracy. But how to do that while 
preserving customer privacy and anonymization and user engagement are challenging issues.

@@

\begin{itemize}
	\item Providing advice on shelters during hurricanes, heat emergencies, and winter blizzards
	\item Handling countless inquires during the COVID pandemic 
	(testing centers, vaccine sites, mobile vaccine clinics, etc.)
	\item Enforcement of standards between landlords and tenants
	\item Re-allocation of NYC's yellow taxi routes based upon an insightful analysis 
	conducted by the Taxi \& Limousine Commission (TLC)
	\item Improved responsiveness for City Agencies with regards to direct 
	client-facing responsibilities such as Parks \& Rec, Dept of Transportation, DSNY, etc.
	\item Increased 311 SRs related to Homelessness resulted in the launch 
	of the City's Homeless Outreach \& Mobile Engagement Street Action Teams (HOME-STAT)
	which includes street canvassers and increases in homeless housing facilities.
	\item A significant increase in graffiti complains in 2017 led directly to 
	the `Graffiti-Free NYC'' campaigns and increased funding for removal of graffiti in public spaces.
	\item Enhanced geospatial data regarding the location of incidents, 
	the responding Agency, and actions taken across a myriad of complaint\_types
	\item Establishment of Agency SLAs in regard to certain complaint types, 
	such as residential noise, homelessness person assistance, street light outages, 
	rat sightings, and other neighborhood quality of life issues.
\end{itemize}

The NYC 311 service exemplifies the dynamic nature of urban data
management and its critical role in modern governance. In addition it provides an
invaluable resource for data scientists to practice data science activities, including data curation.

We extracted 311 SR data for the period of CY 2022-2023, a dataset consisting of
6.4 million records. That dataset serves as the primary source for the analysis effort. 
This is a large dataset which enables observation of some rarely 
seen events, data inaccuracies, and logical inconsistencies. It 
provides enough data to examine trends and data outlier analysis. 
We queried the data directly from the NYC Open Data Portal and then 
exported the data in CSV format for analysis using custom written R programs.

Additionally, we extracted two longer-term datasets, a 5-year look (2019-2023) 
and a 10-year look (2014-2023) ; some 14 and 27 million rows respectively. 
These larger datasets afford an opportunity for long-term trend observations, 
confirming and supporting the analysis using the 2-year dataset. The 2022-2023 dataset
was used exclusively for the detailed, field-by-field level analysis 
which comprises the majority of this effort.

\section{10-year Analysis \& Trends} \label{sec:trends}
Since the NYC 311 System went live in 2003, the usage has grown at an annual
rate of 4.6\%. As part of this study, we also looked at a 10-year period (2014-2023) to
observe system performance over a longer time-frame . During that time-frame, 
the number of 311 SRs grew 50\%. The summer  and early fall months 
make up all of the Top 5 SR counts by day. The all-time daily 
high for SRs, 24,415 SRs (a full 9.8 \textsigma's from the 10-year mean), 
occurred on Tuesday, 2020-08-04. The 10-year low count occurs 
on Christmas Day, 2016, a Sunday. Note also that the difference between 
the peak day (24,415) and the minimum day (2965) varies by 8X, creating a 
challenging system performance issue. This timeframe (2014-2023) 
also includes the COVID pandemic, which reached a peak in 
the July-September 2020 timeframe, accompanied by additional 
spikes in the 2\textsuperscript{nd}half of 2021 as well.

@@

\begin{table}[tbp]
    \centering
    \caption{Top 5 and Bottom 5 Days for SRs}
	    \begin{tabular}{@{}lll lll@{}}
	        \toprule
	        \multicolumn{3}{c}{Top 5 Days} & \multicolumn{3}{c}{Bottom 5 Days} \\
	        \cmidrule(r){1-3} \cmidrule(l){4-6}
	        Date & Count & Rank & Date & Count & Rank \\
	        \midrule
	        2020-08-04 & 24,415 & 1 & 2014-12-25 & 2965 & 1 \\
	        2020-08-05 & 19,560 & 2 & 2015-12-25 & 3247 & 2 \\
	        2023-09-29 & 17,962 & 3 & 2014-04-20 & 3287 & 3 \\
	        2020-07-05 & 16,916 & 4 & 2015-12-26 & 3405 & 4 \\
	        2020-06-21 & 15,883 & 5 & 2016-11-06 & 3419 & 5 \\
	        \bottomrule
	    \end{tabular}
    \label{tab:top-10-counts}
\end{table}

This chart shows the 2-year period which is the basis for this analysis, by calendar month. 
Note that almost all the months before the COVID pandemic outbreak in Feb/Mar
of 2020 are ``below average'', while most of the months after that period 
are ``above average''. That level of higher usage of the 311 system 
has continued into the early months of 2024. It is not surprising that 
the lowest SR counts-per-day occur during the Christmas holiday period.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{2-year-trend-monthly.pdf}
  \caption{Monthly SR Counts for 2022-2023}
  \label{fig:2-yr-monthly}
\end{figure}

Two important fields in this analysis effort are the responsible NYC Agency, 
and the type of complaint. Given the size of the City of New 
York government, it is necessary to identify the responsible 
Agency as well as the type of complaint in order to identify 
the responsible party and assist in troubleshooting discrepancies.

When we discovered a data error, we typically observe one of two trends; 
either a single (or a few) Agencies are responsible (such as Dept of Transportation), 
or it is a system-wide issue that follows the general distribution of  SRs across the system. 
Thus we can often identify if it is an Agency specific problem, or a systemic issue.  
(Note: a few key NYC Agencies are not represented in the 311 
dataset. These include the New York City Housing Authority (NYCHA) 
and the Department of Corrections (DOC).)

Although the 311 SRs involve 16 different agencies, there are 
the ``big six'' Agencies that comprise 90\% of the complaints. These 
six are (in order of \% of complaints):

@@

The distribution of complaint\_type skews to a few select complaints 
accompanied by a long tail. This graph shows the top 20 complaint\_type(s) 
comprising 70\% out of a total of 220.

@@

The top 20 complaints contain several of the eight different types of ``Noise'' 
complaints (Residential, Commercial, Street, Helicopter, etc.). These Noise-related
complaints total 1.435 million in the 2022-2023 dataset, a full 22\% of 
all complaint types, which, when consolidated, is the most frequently occurring complaint\_type.
Illegal Parking is second, followed by Heat/Hot Water. Noise complaints 
are handled by NYPD, as is Illegal Parking and Blocked Driveway complaints, 
which contributes significantly to the top ranking of the NYPD as the lead 
responsible Agency, receiving a full 43\% of all SRs. 
Housing, Preservation, \& Development are 2\textsuperscript{nd}, with 21\%

For  non-NYC residents, HPD is the City agency that manages the 
177,569 NYC public housing units as well as monitoring
all City rental and leased apartment units. HPD collectively serves 
528,105 people, a population larger than Atlanta or Miami. 
Hence the large number of Heat/Hot Water and Unsanitary complaints 
handled by HPD.

@@

There are also some curious and unusual Service Requests in the 311 data: 
tanning, tattooing, trans fat, unsanitary pigeon condition, illegal animal 
kept as pet, harboring bees/wasps, and the authors' favorite
radioactive material. New Yorkers certainly live interesting lives.

@@

\begin{itemize}
	\item Data validation - this effort can span a number of criteria
	\begin{itemize}
		\item Mandatory fields: Certain data fields cannot be empty.
		\item Data-types: Certain fields must be of the correct type, e.g. numeric, 
		character, date, 5 numeric digits, etc. Typically these are identified 
		in a Data Dictionary.
		\item Domain adherence:  Many data fields must adhere to a specific 
		domain of values, e.g. statuses, state names, zip codes, gender. 
		This includes data fields that are restricted to certain ranges, 
		such as longitudes and latitudes.
	\end{itemize}   
	\item Structural errors to include naming conventions, 
	extra fields not in the Data Dictionary, or inconsistent data entry.
	For example intermixing blanks, spaces, NA, N/A, and \textless{}NA\textgreater{} 
	all to indicate the absence of data.
	\item Redundant, unnecessary, or irrelevant fields
	\item Logical inconsistencies such as related fields that violate the 
	nature of that relationship, e.g. a ``due date'' that is before the ``created date''
	\item Data standardization. Often free-form entry fields can suffer 
	from inconsistent data structures. Efforts to standardize these fields 
	can greatly assist analysis.
	\item Accuracy and precision, which of course are not the same thing 
\end{itemize}

This analysis effort is to identify the presence of such errors; not to correct them. 
That effort would be undertaken only after an investigation as to the why
and how such errors came about, and a discussion as to whether or not it 
is even an ``error''. Typically this requires discussions with subject matter 
experts (SMEs) and liaison within the various NYC Agency open data 
coordinators who can help identify the underlying cause for the error. 
For example, an invalid date field that is provided to the NYC Open Data 
data lake via an Agency integration could be the result of selecting 
the wrong field from the source Agency's system, e.g. 
due\_date instead of closed\_date. In other cases, without a subject matter
expert, it is not possible to actually determine if the data is inaccurate or not. 
There are many empty values in the taxi\_company\_borough field. Are those
entries missing or does the entry in that field only apply to a 
very small number of complaint\_type(s) or circumstances. Only a SME
from the Taxi \& Limo Commission (TLC) can provide that answer. 
The R code used here to analyze the 311 Service Request dataset identifies
both obvious inaccuracies as well as potential errors requiring further 
investigation. It is not a conclusive list of errors, but rather mostly identifies
``a place to look'' for further evaluation and potential action.

This paper will focus on a set of specific data cleansing areas:

\begin{itemize}
	\item Examining structural issues within the data
	\item Validating data type correctness
	\item Identifying missing, blank, or N/A data
	\item Identifying invalid values
	\item Exposing logical inconsistencies \& inconsistent or 
	unusual patterns in the data
	\item Accuracy and precision issues
	\item Identifying potential redundant or irrelevant data  
\end{itemize}

\section{Structural Issues}\label{sec:structural}
Structural issues in the context of data cleansing refers to issues 
related to the how data is organized, formatted, or structured within a dataset. Structural issues 
can make it difficult to analyze the data effectively. Some common 
structural concerns in this 311 SR 2022-2023 dataset include:

\begin{itemize}
	\item Data structure (data fields, columns, formats, data types, etc.) 
	not corresponding to the Data Dictionary
	\item Verification of correct data types (numeric, character, geospactial, dates, etc.)
	\item Embedded or combined values: data elements that 
	contain multiple pieces of information 
\end{itemize}

Here are some characteristics of the 311 SR data set:

\begin{itemize}
	\item There are 47 columns of data for each row, exportable as a CSV file.
	\item There are four date fields (created, closed, updated, due).
	\item There are three borough fields; two of which we believe to be duplicates.
	\item Two zip code fields, but not duplicates
	\item Seven street fields; one pair of which we believe to be duplicates
	\item Agency name and Agency abbreviation
	\item Two Police Precinct fields; not duplicates
	\item In addition to street/city, there are three other location fields: 
	lat/long, NY State plane, and Block \#
	\item incident\_address and street\_name, e.g. incident\_address, 
	25 Grymes Hill Road and street\_name (Grymes Hill Road)
	\item One free-form text field, resolution\_description, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

\textbf{Issue: Fields not in the Data Dictionary} The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
 identifies 41 data columns (fields) along with related information 
 for each column. However, when you download the data or 
 explore it via the online portal, you immediately note that there 
 are 47 columns (fields). The additional six fields are not addressed in the 
 Data Dictionary, but do and show up on the Column Manager widget
 as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer 
 what the these @computed\_fields are, but it is not possible to know 
 for certain. These six fields are:

@@

Here is a screenshot of the NYC Open Data Portal ``Column Manager'' 
page showing those columns:

\begin{figure}[tbp]
 	\centering
 	\includegraphics[width = .65\textwidth] {computed_fields_jpg.jpg}	
	\caption{Screenshot of Data Portal with ``@computed'' Columns}  
	\label{fig:computed-columns}
\end{figure}

This, of course, raises several questions:

\begin{itemize}
	\item How are these fields computed? Using what source data? 
	Using what computational method?
	\item How can these fields be verified or cross-checked?
	\item Why are there two fields that appear redundant ``police\_precinct' 
	and ``police\_precincts''? Are these redundant?
	\item What are ``borough boundaries''? They appear to be borough names. 
	Why are they referred to as ``boundaries''?
	\item What are ``community districts''? Is this the same as the
	 ``community\_board(s)'' that are part of the NYC government?
	\item Why are not these fields included in the official 311 SR Data Dictionary?
	\item Can these fields be reliably used for analytical purposes? 
\end{itemize}

The last bullet point is the key here. Are these fields reliable and accurate, 
and can they be used for subsequent analysis? Unfortunately,
as of this date, those questions have not been fully addressed by the 
NYC Open Data Team. This paper will further address that issue by
exploring the accuracy, validity, and candidacy for subsequent analytical usage.

\section{Validating data types}\label{sec:datatypes}
Fortunately, both the Data Dictionary and the portal column manager 
do a good job of indicating the data type. As indicated by 
the red underline in Figure \ref{fig:computed-columns}, there 
is a small icon next to the field name which indicates the 
data type of the field. The stylized capital ``T'' indicates ``text'', the ``\#'' sign indicates 
``numeric'', a calendar icon indicates a ``date'' field, and the pin on a 
map symbol indicates a ``geospatial'' field. The following fields were 
checked for their specified data types:

@@

All data fields were subjected to a test of ``correctness'' 
according to their ``type'' as specified
in the Data Dictionary. Of the nearly 300 million data fields to check 
(6.5 million rows, 47 fields per row), there were only two values that failed 
the test for appropriate data type; those being two text entries in the 
incident\_zip field (``na'', and ``N/A''). All date fields were dates. 
All numbers were numeric. All text fields were character. 
Data typing appears to not be an issue with this dataset.

\section{Blank \& N/A data}\label{sec:blanks}
Understanding the absence of data by field is an important factor 
when undertaking analysis. For example, if you wanted to see if the SRs were
closed before or after their due\_date, you would be challenged 
as 99.57\% of the due\_date field is blank. (Although with such a large dataset
that still leaves 24,411 values, all of which are ``N/A'. Which calls 
into question...are these missing values, or is the concept of a
due\_date simply not applicable for the vast majority of SRs?

When counting the various fields for blank or N/A values, the fields appear 
to divide into three groups: \textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}. he Mostly Empty category ranges from 93-99.9\% blank. 
It includes such fields as taxi\_company due\_date, pickup\_location, and 
landmark. The Partially Empty includes such fields as location\_type, borough, 
and cross\_street. And the Few/None Empty includes created\_date, 
complaint\_type, agency, and status. In some cases it may make 
sense to inquiry as to why some fields are frequently blank. Is the data 
difficult to capture? Does it pertain to only a small set of 
complaint\_type(s) or Agencies? And if the data is truly that sparse, 
does it make sense to collect it? Is it legacy data? For example, 
one would expect that every complaint\_type that has a geographical 
component, such as an address, would also have a corresponding 
police\_precinct, community\_board, and community\_district since these
are all-inclusive throughout the five Boroughs of NYC. Addtionally, 
that it would have a lat/long. And generally that is the case, but these fields 
do have small number, but nonetheless significant, of empty entries.

>>>>

\begin{itemize}
	\item NYC restaurant health inspection violations
	\item Mapping of car crashes involving pedestrians
	\item Visualization of NYC High School and College enrollment
	\item An interactive dashboard to filter and view charges for jail inmates
	\item Location of City-wide free Internet access points
	\item ...and 220 other complaint types. 
\end{itemize}

@@

The paper is organized as follows:
Section~\ref{sec:history} offers a brief review of the history of the 311 system. In 
Section~\ref{sec:trends} we take a look at long-term trends presented 
via a 10-year analysis. Section~\ref{sec:issues} offers a 
general discussion of data cleansing issues impacting data 
quality and curation efficiency. Section~\ref{sec:structural} examines
the technical dimensions of the dataset, the structural issues. 
In Section~\ref{sec:datatypes} we examine the data fields for compliance 
with the Data Dictionary. Section~\ref{sec:blanks} looks at the 
data fields as regards missing, blank, or N/A entries. Section~\ref{sec:domain} 
explores how data fields comply with a domain of legal 
or acceptable values. Section~\ref{sec:inconsistencies} deals 
with the important issues surrounding logical inconsistencies 
and concerning patterns in the data. Section~\ref{sec:precision} 
explores the issue of precision versus accuracy. Section~\ref{sec:duplicates}
identifies duplicate and redundant data fields. Section~\ref{sec:improvements} provides 
actionable suggestions for mitigating or resolving identified issues. 
Following this, Section~\ref{sec:protocol} outlines a series of general 
principles for the release of open data, drawing from our findings. The 
paper concludes with a discussion and recommendations in 
Section~\ref{sec:discussion}, encapsulating the key insights and 
implications of our research.

A note on naming conventions: dataset field names are named 
using ``snake\_case'' where each word in the variable name 
is written in lowercase, and words are separated by 
underscores, e.g. created\_date, complaint\_type, zip\_code, etc..

\section{NYC 311 System History} 
\label{sec:history}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. The evolution of this system can be traced from
its initial implementation as a simple inquiry channel via telephone to a
comprehensive data management system that handles millions of requests
annually. Each inquiry is logged as a complaint, a Service Request (SR), 
in the 311 system. Key milestones in the 311 system include:

\begin{itemize}
 	\item 2003 - NYC 311 system goes live with a phone-based call center only. 
 	The ``311'' phone number replaces a myriad of numbers to 
 	call for each individual City Agency.
   	\item 2009 - 311 online \& mobile apps launched. Today 60\% of all 
   	Service Requests come from mobile \& online channels.
      \item 2020 - In August 2020, driven by the COVID pandemic, monthly 311 
      SRs hit an all-time high of 348,463. Seven of the Top 10 all-time busiest 
      days occur in 2020.
      \item 2021 - 311 System is expanded to include the MTA's city subway system, 
      the largest expansion since its inception.
   	\item 2023 - Record high year for 311 with 3.23 million Service Requests.
\end{itemize}

Today, the NYC 311 data system is a robust platform that manages a
vast array of urban living-related inquiries. The system handles over 3 million service 
requests (SRs) per year, encompassing such complaints as street noise, 
illegal parking, heat/hot water, abandoned vehicles,
and unsanitary conditions. The data is a valuable resource for City
administration and policy-making. The 311 data is publicly accessible through the 
NYC Open Data Portal which provides access to all the data 
sets (approx. 2700) and provides a convenient method for querying, 
grouping, aggregating, geo-mapping, visualization, and exporting 
results. This open data  enables not only governmental transparency 
but empowers researchers, civic developers, and the general public.

Despite its success, the 311 system faces challenges such as:

\begin{itemize}
	\item Data timeliness, accuracy, and consistency
	\item Difficulties correlating data over long time periods, e.g. over a 10-year 
	period, e.g. Agency name changes, new complaint types.
	\item Data anonymization and handling of Personally identifiable information (PII)
	\item Integration with stand-alone systems at selected NYC Agencies
	\item Managing API usage, authentication, and load for 3\textsuperscript{rd} 
	party users, of which there are many
\end{itemize}

These and other challenges are addressed by the various 
Agency open data managers as well as the 
\href{https://www.nyc.gov/content/oti/pages/}
{NYC Office of Technology and Innovation (OTI)} which provides the 
technology support for the open data system. Typically, the data 
will be exported and processed in an external program. The authors found 
this to be the case with the large 311 dataset examined.

@@

\begin{itemize}
	\item Providing advice on shelters during hurricanes, heat emergencies, and winter blizzards
	\item Handling countless inquires during the COVID pandemic 
	(testing centers, vaccine sites, mobile vaccine clinics, etc.)
	\item Enforcement of standards between landlords and tenants
	\item Re-allocation of NYC's yellow taxi routes based upon an insightful analysis 
	conducted by the Taxi \& Limousine Commission (TLC)
	\item Improved responsiveness for City Agencies with regards to direct 
	client-facing responsibilities such as Parks \& Rec, Dept of Transportation, DSNY, etc.
	\item Increased 311 SRs related to Homelessness resulted in the launch 
	of the City's Homeless Outreach \& Mobile Engagement Street Action Teams (HOME-STAT)
	which includes street canvassers and increases in homeless housing facilities.
\end{itemize}

We extracted 311 SR data for the period of CY 2022-2023, a dataset of
6.4 million SR records. This is a large dataset which enables 
observation of some rarely seen events, data inaccuracies, 
and logical inconsistencies. It provides enough data to 
examine trends and data outlier analysis. We queried the data 
directly from the NYC Open Data Portal and exported the 
data for analysis with custom written R programs.

\section{10-year Analysis \& Trends} 
\label{sec:trends}
As part of this study, we also looked at a 10-year period (2014-2023) to
observe system performance over a longer time-frame . During that time-frame, 
the number of 311 SRs grew by 50\%. The summer and early fall months 
make up all of the Top 5 SR counts by day. The all-time daily 
high for SRs, 24,415 SRs (a full 9.8 \textsigma's from the 10-year mean), 
occurred on Tuesday, 2020-08-04. The 10-year low count occurs 
on Christmas Day, 2016, a Sunday.

@@

\begin{table}[tbp]
    \centering
    \caption{Top 5 and Bottom 5 Days for SRs}
	    \begin{tabular}{@{}lll lll@{}}
	        \toprule
	        \multicolumn{2}{c}{Top 5 Days} & \multicolumn{2}{c}{Bottom 5 Days} \\
	        \cmidrule(r){1-2} \cmidrule(l){3-4}
	        Date & Count & Date & Count\\
	        \midrule
	        2020-08-04 & 24,415 & 2014-12-25 & 2965\\
	        2020-08-05 & 19,560 & 2015-12-25 & 3247\\
	        2023-09-29 & 17,962 & 2014-04-20 & 3287\\
	        2020-07-05 & 16,916 & 2015-12-26 & 3405\\
	        2020-06-21 & 15,883 & 2016-11-06 & 3419\\
	        \bottomrule
	    \end{tabular}
    \label{tab:top-10-counts}
\end{table}

Two important fields in this analysis effort are the responsible NYC Agency, 
and the type of complaint. Given the size of the City of New 
York government, it is necessary to identify the responsible 
Agency as well as the type of complaint in order to identify 
the responsible party and assist in troubleshooting discrepancies.
Thus we can often identify if it is an Agency specific problem, 
or a systemic issue. Although 311 SRs involve 16 different agencies, 
there are six Agencies that comprise 90\% of the complaints.

@@

The distribution of complaint\_type skews to a few select complaints 
accompanied by a long tail. This graph shows the top 20 complaint\_type(s) 
comprising 70\% out of a total of 220 different types.

@@

The top 20 complaints contain several of the eight different types of ``Noise'' 
complaints (Residential, Commercial, Street, Helicopter, etc.). These Noise-related
complaints total a full 22\% of all complaint types, which, when consolidated, is the most frequently occurring complaint\_type.

@@

@@

\begin{itemize}
	\item Data validation - this effort can span a number of criteria
	\begin{itemize}
		\item Mandatory fields: Certain data fields cannot be empty.
		\item Data-types: Certain fields must be of the correct type, e.g. numeric, 
		character, date, 5 numeric digits, etc. Typically these are specified 
		in a Data Dictionary.
		\item Domain compliance:  Many data fields must adhere to a specific 
		domain of values, e.g. statuses, state names, zip codes, gender. 
	\end{itemize}   
	\item Structural errors to include naming conventions, 
	 fields not in the Data Dictionary, or inconsistent data entry. For example 
	 intermixing blanks, spaces, NA, N/A, and \textless{}NA\textgreater{} 
	 all to indicate the absence of data.
	\item Redundant, unnecessary, or irrelevant fields
	\item Logical inconsistencies such as related fields that violate the 
	nature of that relationship, e.g. a ``due date'' that is before the ``created date''
	\item Accuracy versus precision
\end{itemize}

This analysis effort is to identify the presence of such errors; not to correct them. 
That effort would be undertaken only after an investigation as to the why
and how such errors came about, and a discussion as to whether or not it 
is even an ``error''.

\section{Structural Issues}
\label{sec:structural}
Structural issues in the context of data cleansing refers to issues 
related to the how data is organized, formatted, or structured within a 
dataset. Structural issues can make it difficult to analyze the 
data effectively. Here are some characteristics of the 311 SR data set:

\begin{itemize}
	\item There are 47 columns of data for each row, exportable as a CSV file.
	\item There are four date fields (created, closed, updated, due).
	\item There are three borough fields; two of which appear to be duplicates.
	\item Two zip code fields, but not duplicates
	\item Seven street fields; one pair of which appear to be duplicates
	\item Two Police Precinct fields; not duplicates
	\item In addition to incident\_address, there are five additional location fields: 
	lat/long, street\_name, landmark, NY State plane, and Block \#
	\item One free-form text field, resolution\_description, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

\textbf{Issue: Fields not in the Data Dictionary} The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
 identifies 41 data columns (fields) along with related information 
 for each column. However, when you download the data or 
 explore it via the online portal, you immediately note that there 
 are 47 columns (fields). The additional six fields are not addressed in the 
 Data Dictionary, but do and show up on the Column Manager widget
 as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer 
 what the these @computed\_fields are and how they are determined, but 
 it is not possible to know for certain. Most importantly, can these fields 
 be reliably used for analytical purposes?  These six fields are:

@@

\section{Validating data types}
\label{sec:datatypes}
Fortunately, both the Data Dictionary and the portal column manager 
do a good job of indicating the data type. The following fields were 
checked for their specified data types:

@@

\section{Blank \& N/A data}
\label{sec:blanks}
Understanding the absence of data by field is an important factor 
when undertaking analysis. For example, if you wanted to see if the SRs were
closed before or after their due\_date, you would be challenged 
as 99.57\% of the due\_date field is blank. Which calls into 
question...are these missing values, or is the concept of a 
due\_date simply not applicable for the vast majority of SRs?

When counting the various fields for blank or N/A values, the fields appear 
to divide into three groups: \textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}. he Mostly Empty category ranges from 93-99.9\% blank. 
It includes such fields as taxi\_company due\_date, pickup\_location, and 
landmark. The Partially Empty includes such fields as location\_type, borough, 
and cross\_street. And the Few/None Empty includes created\_date, 
complaint\_type, agency, and status. In some cases it may make 
sense to inquiry as to why some fields are frequently blank. Here is a graphic 
depiction of total empty (blank \& N/As) for each fields. You can see 
the natural grouping into the Most, Some, and Few categories 
in the stair-type visualization.
